Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-24 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches have left several critical issues
unresolved. This study examines two such problems that can compromise
the reliability and validity of CJ's trait estimates. The first issue
concerns the overreliance on Thurstone's Case V assumptions in the
statistical analysis of CJ data. The second involves the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing.

The study has five main sections. It begins with a brief overview of
Thurstone's theory in Section~\ref{sec-thurstone_theory}, followed by a
detailed examination of these issues in Section~\ref{sec-theory-issues}.
In Section~\ref{sec-theory}, the study introduces a theoretical model
for CJ that builds upon Thurstone's theory, along with its statistical
translation, designed to address both issues simultaneously. Finally,
Section~\ref{sec-discussion} discusses the concepts developed in the
study, their implications, and avenues for future exploration, while
Section~\ref{sec-conclusion} presents the conclusions.

@@

This section examines the two critical issues in CJ's theory that can
compromise the reliability and validity of CJ's trait estimates. In
Section~\ref{sec-theory-issue1}, the study examines the first issue: the
overreliance on Thurstone's Case V assumptions in the statistical
analysis of CJ data. In Section~\ref{sec-theory-issue2}, the study
explores the second issue: the apparent disconnect between CJ's approach
to trait measurement and hypothesis testing.

@@

\section{Updating the theoretical and statistical
model}\label{sec-theory}

The objective of this section is twofold. First, it aims to refine CJ's
theoretical model using the \emph{structural approach} to causal
inference. This approach, as outlined in
Section~\ref{sec-theory-theoretical}, defines a model that integrates
the core principles of Thurstone's theory and key assessment design
features relevant to CJ experiments. Second,
Section~\ref{sec-theory-statistics} uses \emph{Bayesian inference
methods} to convert these theoretical and practical elements into a
statistical model that supports the analysis of paired comparison data.

\subsection{The theoretical model}\label{sec-theory-theoretical}

The \emph{structural approach} to causal inference provides a formal
framework to identify causes and estimate their effects using data. The
approach uses structural causal models (SCMs) and directed acyclic
graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system. Essentially, SCMs and DAGs serve as \emph{conceptual
(theoretical) models} that facilitate valid causal inference. They
achieve this by helping researchers to determine which statistical
models can identify an estimand, assuming the depicted causal structure
is correct \citep{McElreath_2020}. \emph{Identification} determines
whether an estimator can accurately compute an estimand based on its
assumptions, regardless of random variability
\citep[pp.~4]{Schuessler_et_al_2023}.

SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence between
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering offering the
advantage of highlighting potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

Therefore, this section leverages SCMs and DAGs to represent the
conceptual population of CJ experiments, referred to as \emph{the
population model} in Section~\ref{sec-theory-theoretical_P}, and later
integrates various assessment design features relevant to CJ
experiments. This process leads to the development of \emph{the
sample-comparison model} in Section~\ref{sec-theory-theoretical_SC}.

\subsubsection{The population model}\label{sec-theory-theoretical_P}

@@

\includegraphics[width=0.23\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_01.png}

\caption{\label{fig-CJ_TM_01}}

\includegraphics[width=0.32\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_02.png}

\caption{\label{fig-CJ_TM_02}}

\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}

\caption{\label{fig-CJ_TM_03}}

@@

\includegraphics[width=0.59\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_06.png}

\caption{\label{fig-CJ_TM_06}}

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_07.png}

\caption{\label{fig-CJ_TM_07}}

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_08.png}

\caption{\label{fig-CJ_TM_08}}

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}

\caption{\label{fig-CJ_TM_09}}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}

\caption{\label{fig-CJ_TM_11}}

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}

\caption{\label{fig-CJ_TM_13}}

@@

\subsection{From theory to statistics}\label{sec-theory-statistics}

Bayesian inference procedures offer three key advantages. First, they
are well-suited to handling complex and overparameterized models,
enabling researchers to estimate models where the number of parameters
exceeds the number of observations for estimation
\citep{Baker_1998, Kim_et_al_1999}. Second, they allow researchers to
incorporate prior information, which helps constrain parameters within
specified bounds. This capability addresses challenges such as
non-convergence or improper parameter estimation that often arise in
complex models when analyzed with frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. Finally, Bayesian methods
are particularly effective at drawing inferences from small sample
sizes, where relying on the asymptotic properties of frequentist
approaches may not be justified
\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.

@@

Researchers would then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute an estimand based on its assumptions, regardless of random
variability \citep[pp.~4]{Schuessler_et_al_2023}. An \emph{estimator} is
a method or function that transforms data into an estimate
\citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand, and they are derived through
\emph{estimation}, referring to the process of integrating data with an
estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
provides a visual representation of the process of transitioning from
estimands to estimates.

@@

While numerous estimators can approximate an estimand, researchers
prioritize estimators with desirable identification properties that
ensure the accuracy of estimates. For instance, the Z-test is an
estimator known for its effectiveness when comparing groups'
proportions, yielding accurate estimates when the underlying assumptions
of the statistic are met \citep{Kanji_2006}. If this is the case, the
Z-test is expressed as a signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i} \:|\:T_{i}=1]\) and \(E[Y_{i} \:|\:T_{i}=2]\), respectively.
The noise, represented by \(\hat{s}_{p}\), is defined as the unpooled
sample variability observed between the two groups.

However, researchers often seek to uncover the mechanisms underlying
specific data and establish causal relationships rather than simply
estimate associations. In the example, researchers can interpret the
associational estimand in Equation~\ref{eq-group_diff} and the
associational estimate represented by the Z-statistic as causal. This
interpretation is valid as long as the data satisfy the assumptions of
the Z-test and are collected through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key properties, such as
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from any other variable \(X\)
\citep{Morgan_et_al_2014, Neal_2020}.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that facilitate valid causal inference. They achieve this by
helping researchers to determine which statistical models can identify
an estimand (ACE, CACE, or other), assuming the depicted causal
structure is correct \citep{McElreath_2020}. Figure~\ref{fig-IEflow}
shows the role of theoretical models in the inference process.

>>>>

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches have left several critical issues
unresolved. This study addresses two such issues that can compromise the
reliability and validity of CJ's trait estimates. The first concerns the
overreliance on Thurstone's Case V assumptions in the statistical
analysis of CJ data. The second focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

The study is organized into six main sections. It begins with an
overview of Thurstone's theory in Section~\ref{sec-thurstone_theory},
followed by a detailed examination of these issues in
Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
study develops a theoretical model for CJ that simultaneously addresses
these issues. The model integrates Thurstone's core principles with key
assessment design features relevant to CJ experiments, such as the
selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a statistical model that supports the analysis of pairwise
comparison data. Finally, Section~\ref{sec-discussion} discusses the
concepts developed in the study, their implications, and avenues for
future exploration, while Section~\ref{sec-conclusion} summarizes the
findings.

@@

This section explores the two critical issues in CJ's theory that have
the potential to compromise the reliability and validity of CJ's trait
estimates. Section~\ref{sec-theory-issue1} inspects the overreliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

\section{Updating the theoretical model}\label{sec-theoretical}

\subsection{The population model}\label{sec-theory-theoretical_P}

@@

@@

@@

@@

@@

\section{From theory to statistics}\label{sec-statistical}

@@

Ideally, researchers then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute the estimand based solely on its assumptions, regardless of
random variability \citep[pp.~4]{Schuessler_et_al_2023}. Identification
is crucial because if an estimator cannot reliably compute an estimand
using ``infinite'' and error-free data, it is impossible to gain
meaningful insights about that estimand using finite data
\citep[pp.~5]{Schuessler_et_al_2023}. An \emph{estimator} refers to a
method or function that transforms data into an estimate
\citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand derived through the process of
\emph{estimation}, which integrates data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020}, shown in Figure~\ref{fig-IEflow},
visually represents the transition from estimands to estimates.

@@

Consequently, although many estimators can approximate an estimand,
researchers prioritize those with desirable identification properties to
ensure accurate and reliable estimates. For instance, the Z-test is a
widely used estimator for comparing group proportions, providing
accurate estimates when its underlying assumptions are satisfied
\citep{Kanji_2006}. Moreover, researchers can interpret the Z-test
estimates as causal, provided they collect the data through a randomized
experiment.

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key identification
properties, such as common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from any other variable \(X\)
\citep{Morgan_et_al_2014, Neal_2020}.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, these tools serve as \emph{conceptual (theoretical) models}
on which identification analysis rests
\citep[pp.~4]{Schuessler_et_al_2023}. Specifically, they guide
researchers in determining which statistical models can identify an
estimand (ACE, CACE, or other), assuming the depicted causal structure
is correct \citep{McElreath_2020}, thus enabling valid causal inference.
Figure~\ref{fig-IEflow} shows the role of theoretical models in the
inference process.
